---
title: "Linear Regression"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

defaultW <- getOption("warn") 
options(warn = -1) 
```

### Linear Regression on GDP dataset

The site Gapminder compiled a dataset with population, life expectancy and GDP per capita of 142 countries over 12 years. Let us analyze the data.

Importing data:

```{r import_data_2}
gdp = read.csv2("C:\\Users\\dougl\\Documents\\DOUGLAS\\Data_Scientist\\data\\gdp_gapminder.csv", sep = ",", header = TRUE)
```

Checking the data structure:

```{r str_data_2}
str(gdp)

library(dplyr)

gdp = gdp %>% mutate(pop = as.double(pop), expLife = as.integer(expLife), gdpPercap = as.numeric(gdpPercap))

```

Data summary:

```{r summary_data_2}
summary(gdp)
```

Tables with absolute and relative frequency of observations per continent:

    i) Absolute frequency

```{r absolut_data_2}
cont_abs <- data.frame(table(gdp$continent))
cont_abs
```

    ii) Relative frequency

```{r relative_data_2}
cont_rel <- data.frame(round(table(gdp$continent)/length(gdp$continent),2)*100)
cont_rel
```

Bar chart with absolute frequency of continents:

```{r abs_bar_data_2}
library(ggplot2)
library(ggpubr)
ggplot(gdp, aes(x=continent, fill=continent)) + geom_bar() + ylab("Frequency")
```

Graph correlating GDP per capita with life expectancy:

```{r rel_pib_exp_data_2}
g <- ggplot(data = gdp, aes(x = expLife, y = gdpPercap, color = continent)) + geom_point() + labs(x = "Life Expectancy", y = "GDP per Capita")
g
```

Creating 2 new columns in the same dataset with log of GDP per capita and log of Life expectancy:

```{r logs_data_2}
library(plm)

lgdpPercap <- log(gdp$gdpPercap)
lexpLife <- log(gdp$expLife)
gdp["lgdpPercap"] <- lgdpPercap
gdp["lexpLife"] <- lexpLife
str(gdp)
```

Graph correlating these 2 new attributes:

```{r rel_logs_data_2}
g <- ggplot(data = gdp, aes(x = lexpLife, y = lgdpPercap, color = continent)) + geom_point() + labs(x = "Life Expectancy", y = "GDP per Capita")
g
```

Adjusting a linear model to data by using these 2 new attributes and being lexpLife the dependent variable:

```{r model_logs_data_2}
X <- gdp$lgdpPercap
Y <- gdp$lexpLife

model <- lm(Y ~ X, gdp)
model
```

Performing the regression analysis:

    i)   Diagnosing graphics

```{r graphs_logs_data_2}
g <- ggplot(data = gdp, aes(x = X, y = Y)) + geom_point() + xlab("GDP per Capita") + ylab("Life Expectancy")
g

par(mfrow=c(2,2))
plot(model)
```

As per the scatter plots we can see a correlation between life expectancy and GDP per capita but there is no clear linearity.

In the residual x fitted values graph there is a certain residual concentration and a decrease of the same as the adjusted values increase. We can also see that residuals do not follow a normal distribution on Normal Q-Q graph.

    ii) Evaluate model summary

```{r summary_logs_data_2}
summary(model)
```

We can observe that both intercept and independent variables have p-values far below the significance level of 0.05 and hence we can **reject** the null hypothesis that coefficients are equals to zero.

    iii) Explanatory power of the model

The multiple R squared indicates that our model explains 61.3% of the target attribute variance. Hence the model could be enhanced.

```{r histogram_logs_data_2}
model.df <- as.data.frame(residuals(model))
names(model.df) <- c("residuals")
g <- ggplot(data = model.df, aes(residuals)) + 
  geom_histogram(fill="#aabbee",color = "blue",stat = "bin", bins = 5) + 
  theme_pubr() + labs_pubr() + xlab("Residuals") + ylab("Frequency") +  
  theme(plot.caption = element_text(hjust = 0)) + 
  labs(caption = "Source: Elaborated by the author" )
g
```

```{r include=FALSE}
options(warn = defaultW)
```